---
title: "BIOF2014: Statistical Modelling in Bioinformatics"
author: "David J. H. Shih"
format: pdf
---

```{r setup, include=FALSE}
options(warn=-1)
knitr::opts_chunk$set(echo = TRUE, fig.align = "center");
```

## Intended learning outcomes

- Describe the motivations for establishing and applying statistical models


## How is statistical modeling useful?

- We want to model the real world
- Many quantities of interest have uncertainty
- Deterministic models can lead to unrealistic predictions
- Stochastic models account for randomness in real-world phenomenons


## Example of a deterministic system: Newtonian physics

All variables are measurable to arbitrary precision.

### Newton's laws of motion
$$
\begin{aligned}
F = 0 \; \Rightarrow \; \frac{dv}{dt} &= 0 \\
\frac{dv}{dt} m &= F \\
F_{1 \rightarrow 2} &= F_{2 \rightarrow 1}
\end{aligned}
$$


## Example of a stochastic system: quantum mechanics

At the atomic scale, some quantum variables cannot be measured 
precisely at the same time.

### Heisenberg's Uncertainty Principle
For uncertainty in momentum $\Delta p$ and uncertainty in position $\Delta x$,
$$
\Delta p \, \Delta x \ge \frac{h}{4 \pi}, 
$$
where $h$ is Planck's constant.


## A simple deterministic system

Statistics is useful for modeling systems with high uncertainty or with
stochasticity. What about deterministic systems?

Let's consider the following deterministic system.

Let $x \in (0, 1)$ be a quantity of interest
(e.g. position of a particle).

Define a function $f$ as
$$
f(x) \triangleq r x (1 - x) - x
$$
We model the dynamics of $x$ under continuous time $t \ge 0$ by
$$
\frac{dx}{dt} = f(x) .
$$

We simplify the model to discrete time $t \in \{0, 1, 2, \ldots \}$ by
$$
\frac{\Delta x}{\Delta t} = f(x) .
$$
We can simplify further by setting $\Delta t = 1$, giving
$$
x_{t + 1} = x_t + f(x_t) ,
$$
which is equivalent to
$$
x_{t + 1} = g(x_t),
$$
where
$$
g(x) \triangleq r x (1 - x) .
$$

### Iterated function applications

For $t > 1$, let $f_{t+1}(x) = f( f_t(x) )$. Let $f_1(x) = f(x)$.

For example,
$$
\begin{aligned}
x_2 &= f_2(x_0) = f( f(x_0) )          \\
x_3 &= f_3(x_0) = f( f( f(x_0) ) )     \\
x_4 &= f_4(x_0) = f( f( f( f(x_0)  ) ) \\
&\ldots \\
x_t &= f_t(x_0) = f( \ldots f( f(x_0) ) \ldots )
\end{aligned}
$$


Also,
$$
x_\infty \triangleq \lim_{t \rightarrow \infty} x_t
$$



## Simulation under deterministic model

```{r}
library(ggplot2)

# x \in (0, 1)
logistic_map <- function(x, r) {
  r * x * (1 - x)
}

trace_plot <- function(g, x0, T=100) {
  xs <- numeric(T);
  xs[1] <- x0;
  for (i in 2:T) {
    xs[i] <- g(xs[i-1]);
  }
  qplot(1:T, xs, geom="line") + theme_classic() +
    xlab("t") + ylab(expression(x[t])) + ylim(0, 1)
}
```

```{r, fig.width=6, fig.height=1.5}
g <- function(x) logistic_map(x, r=1.0);
trace_plot(g, x0=0.3)
```

```{r, fig.width=6, fig.height=1.5}
trace_plot(g, x0=0.6)
```

```{r, fig.width=6, fig.height=1.5}
g <- function(x) logistic_map(x, r=2.0);
trace_plot(g, x0=0.3)
```

```{r, fig.width=6, fig.height=1.5}
trace_plot(g, x0=0.6)
```

```{r, fig.width=6, fig.height=1.5}
g <- function(x) logistic_map(x, r=3.0);
trace_plot(g, x0=0.3)
```

```{r, fig.width=6, fig.height=1.5}
trace_plot(g, x0=0.6)
```

```{r, fig.width=6, fig.height=1.5}
g <- function(x) logistic_map(x, r=4.0);
trace_plot(g, x0=0.3)
```

```{r, fig.width=6, fig.height=1.5}
trace_plot(g, x0=0.6)
```

```{r, echo=TRUE}
iterate <- function(g, x0, T=100) {
  x <- x0;
  for (i in 1:T) {
    x <- g(x);
  }
  x
}

simulate <- function(g, x0) {
  data.frame(
    x.0 = x0,
    x.inf = vapply(x0,
      function(x0) iterate(g, x0), 0
    )
  )
}

```

```{r, echo=TRUE}
predict_plot <- function(g, x0) {
  d <- simulate(g, x0);
  qplot(d$x.0, d$x.inf, geom="point") + 
    theme_classic() +
    xlab(expression(x[0])) +
    ylab(expression(x[infinity])) +
    xlim(0, 1) + ylim(0, 1) + coord_fixed()
}

step <- 0.02;
x0 <- seq(0 + step, 1 - step, by=step);
```

```{r, fig.width=3, fig.height=3, out.width="40%"}
g <- function(x) logistic_map(x, r=2.0);
predict_plot(g, x0)
```

```{r, fig.width=3, fig.height=3, out.width="40%"}
g <- function(x) logistic_map(x, r=4.0);
predict_plot(g, x0)
```

## Chaos

Depending on setting of the parameters, dynamical systems can converge or
exhibit

- periodic behaviour
- *chaotic* behaviour

In the field of differential equations, periodic and chaotic systems 
are *pathological*.


## Probabilistic view of the deterministic system


```{r, fig.width=6, fig.height=1.5, out.width="80%"}
set.seed(1337); N <- 1000;
g <- function(x) logistic_map(x, r=4.0);
x.0 <- rbeta(N, 20, 10); x.inf <- iterate(g, x.0);
qplot(x.0, ylab="count") + theme_classic()
```

```{r, fig.width=6, fig.height=1.5, out.width="80%"}
qplot(x.inf, ylab="count") + theme_classic()
```


## Statistics

Study of uncertainty

- establish statistical models
- estimate parameters of the models
- quantify uncertainty in parameter estimates
- predict new quantities with uncertainty


## Equations

### Deterministic

$$
y = m x + b ,
$$
where $x$ and $y$ are variables; $m$ and $b$ are constants.

\bigskip

### Probabilistic

$$
p \left(Y = y \mid \mu, \tau \right) =
 (2 \pi)^{-\frac{1}{2}} \tau^{\frac{1}{2}} \,
 \exp \left( -\frac{1}{2} \tau (y - \mu)^2 \right)
$$
where $Y$ is a *random* variable and $y$ its *realized value*; 
$\mu$ and $\tau$ are parameters.


## Atomic models

### Bohr model

Electrons travel in different *circular* orbitals around a nucleus.

\bigskip

### Quantum mechanical model

Electron are distributed in different *cloud* orbitals around a nucleus.


## Applications of statistical models

- Demography
- Gambling
- Quantum mechanics
- Biology
- Epidemiology
- Medicine
- Psychology
- Genetics
- Machine learning
- Data science
- Bioinformatics
- ...


## Summary

Albert Einstein: "God does not play dice with the universe."

Niels Bohr: "Einstein, stop telling God what to do."

\bigskip

### Intended learning outcomes

- Describe the motivations for establishing and applying statistical models
