---
title: "Birthday Problem"
author: "BIOF2014"
format: pdf
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, fig.align="center", warning=FALSE);
```

## Problem

What is the probability that at least two people in this classroom share
the same birthday?

## Model

### Assumptions

- 365 days in a year
- Every birthday is equally likely

### Variables

Let $N$ be the number of people in this class.  
Let $x$ be the number of people with the same birthday.

### Task

1. Derive a mathematical equation for $P( x \ge 2 )$.
2. Implement this equation and compute $P( x \ge 2 )$.


## Hint

$$
P( x \ge 2 ) \; = \; P( x > 1 ) \; = \; 1 - P(x = 1) .
$$

\pagebreak

## Simulation

```{r}
set.seed(1234);

# Simulate birthdays and return the number of people who share the same birthday
# N  number of people
# J  number of possible birthdays
simulate <- function(N, J) {
  birthdays <- sample.int(J, N, replace=TRUE);
  counts <- table(birthdays);
  max(counts)
}

N <- 30;
J <- 365;
S <- 1000;  # number of simulation rounds
x <- vapply(1:S, function(s) simulate(N, J), 0);
event <- as.integer(x > 1);
estimate <- mean(event);
se <- sd(event) / sqrt(S);
z <- qnorm(1 - 0.05 / 2);
```

estimate: `r estimate`  
standard error: `r se`  
95% confidence interval: (`r estimate - z * se`, `r estimate + z * se`)  

```{r, echo=FALSE}
p1 <- 1;
for (n in 0:(N-1)) {
  p1 <- p1 * (J - n) / J;
}
```

ground truth: `r 1 - p1`

```{r, fig.width=2, fig.height=2, out.width="25%"}
library(ggplot2)
ggplot(data.frame(x=x), aes(x)) + theme_classic() +
  geom_histogram(binwidth=1) + ylab("count") + ylim(0, S)
```

